---
title: Differential Digestion
author: Soren Heidelbach
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_depth: 2
  
---

# Differential Digestion

```{r setup, setup}
setwd("/home/ubuntu/vol_store/differential_digestion")
pacman::p_load(
  "data.table",
  "magrittr",
  "stringr",
  "ggplot2",
  "ggside",
  "patchwork",
  "gt",
  "gtExtras",
  "knitr",
  "kableExtra",
  "seqinr"
)
knitr::opts_knit$set(
  root.dir = "/home/ubuntu/vol_store/differential_digestion"
)
knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  echo = TRUE,
  fig.path = "analysis/figures",
  out.width = "95%"
)

plot_theme <- theme_bw(
  base_size = 12,
  base_family = "Helvetica"
)


```
  


```{r loadData}
chunk_size_saved  <- 1e3
chunk_size <- 5e3
chunk_file  <- "data/processed/mmlong2_assembly_chunk_depths.tsv"

if (file.exists(chunk_file)) {
  chunks  <- fread(chunk_file, sep = "\t")[
      , chunk := ceiling(chunk * (chunk_size_saved / chunk_size))
    ][
      , .(depth = mean(depth)), by = c("contig", "chunk", "sample")
    ]
} else {
  # Load depth data
  samples <- list.files("data/processed/mappings", full.names = TRUE)
  samples <- samples[grepl('MboI|400', samples)]
  depth <- lapply(
    samples,
    function(sample) {
      depth_file <- paste0(sample, "/depth.txt")
      if (file.exists(depth_file)) {
        dt <- fread(depth_file, header = FALSE, sep = "\t") %>%
          setnames(c("contig", "pos", "depth"))
        dt[, sample := basename(sample)]
        dt
      } else {
        NULL
      }
    }
  )
  depth <- rbindlist(depth)
  depth[
    , chunk := (pos %/% chunk_size_saved) + 1
  ]
  chunks <- depth[
  , .(
    depth = mean(depth)
  ), by = c("contig", "chunk", "sample")
  ]
  fwrite(chunks, "data/processed/mmlong2_assembly_chunk_depths.tsv", sep = "\t", quote = FALSE)
}


contig_to_bin <- fread("mmlong2/tmp/binning/contig_bin.tsv", header = FALSE) %>%
  setnames(c("contig", "bin"))

bins <- fread("mmlong2/results/mmlong2_bins.tsv")[
  , quality := ifelse(Completeness >= 90 & Contamination <= 5, "HQ", "MQ")
  ][
    , Type := "mmlong2"
  ][
    , GC_Content := GC_Content * 100
  ]


chunks <- Reduce(
  function(x, y) {
    shared_columns  <- intersect(names(x), names(y))
    message(paste0(
      "Merging by ",
      paste(shared_columns, collapse = ", ")
    ))
    merge(x, y, by = shared_columns, all = TRUE)
  },
  list(chunks, contig_to_bin, bins)
)

chunks[
    grepl("MboI", sample), sample := "MboI_digest"
  ][
    grepl("400bps", sample), sample := "Nondigest"
  ][
    , bin_char := str_extract(bin, "\\d+$")
  ][
    , bin_num := as.numeric(bin_char)
  ][
    , bin_char := fifelse(is.na(bin_char), "Unbinned", bin_char)
  ][
    , depth_centered := depth - mean(depth), by = c("bin_char")
  ][
    , depth_standard := depth_centered / sd(depth_centered), by = c("bin_char")
  ][
    , depth_log10 := log10(depth)
  ][
    , depth_log10_centered := depth_log10 - mean(depth_log10), by = c("bin_char")
  ][
    , depth_scaled_mean := depth / mean(depth) - 1, by = c("bin_char", "sample")
  ]

chunks_wide  <- dcast(
  chunks,
  bin_num + bin_char + chunk + contig ~ sample,
  value.var = c("depth", "depth_centered", "depth_standard", "depth_log10_centered")
)

get_stats <- function(chunks, col) {
  chunks[
    get("col") != -Inf, list(
      mean = mean(get(col), na.rm = TRUE),
      median = median(get(col), na.rm = TRUE),
      sd = sd(get(col), na.rm = TRUE),
      n = .N,
      q_max = quantile(get(col), probs = 0.75, na.rm = TRUE),
      q_min = quantile(get(col), probs = 0.25, na.rm = TRUE),
      q_75 = quantile(get(col), probs = 0.75, na.rm = TRUE),
      q_25 = quantile(get(col), probs = 0.25, na.rm = TRUE),
      min = min(get(col), na.rm = TRUE),
      max = max(get(col), na.rm = TRUE)
    ), by = c("bin_char", "sample")
  ][
    , c("CI_max", "CI_min", "IQR") := .(
      mean - qt(0.999, n) * sd / sqrt(n),
      mean - qt(0.001, n) * sd / sqrt(n),
      q_75 - q_25
    )
  ]
}
chunks_stats <- get_stats(chunks, "depth")
chunks_stats_scaled_mean <- get_stats(chunks, "depth_scaled_mean")

chunks_sampleVSbin <- data.table::dcast(chunks, bin_char ~ sample, value.var = "depth", fun = list)
chunks_sampleVSbin[
  , p_val := p.adjust(
      wilcox.test(
        log10(Nondigest[[1]] + 1e-300), 
        log10(MboI_digest[[1]] + 1e-300)
      )$p.value,
      method = "bonferroni",
      n = nrow(chunks_sampleVSbin)
    )
  , by = bin_char
]
```

```{r plotParameter}
binary_colors  <- c("#a85959", "#0038b1")
x_axis_order <- chunks[
  sample == "Nondigest", .(median_depth_all = median(depth)), by = bin_char
  ][
    order(median_depth_all, decreasing = TRUE),
    bin_char
  ]
x_axis_order <- c("Unbinned", x_axis_order[x_axis_order != "Unbinned"])
```

## Motivation
To identify if the depletion by digestion is consistant within previously identified bins, to such an degree that differential targeting of digestion is believeable.

1. The first objevtive is show differential depletion of bins, meaning there is some signal

2. The second objective is to show that the depletion is somewhat consistant within bins after digestion

Non parametric statistic is used to reduce the impact of outliers e.g. contig that may or may not be binned corretcly.

## General distrubution of chunks
Bloxplot of coverage of chunks within each bin. Black bar added to show all chunks below coverage 1 (Figure \@ref(chunkBoxplot)).
```{r chunkBoxplot}
ggplot(chunks) +
  aes(x = bin_char, y = depth, color = sample) +
  geom_boxplot(outlier.alpha = 0, outlier.size = 0.1) +
  scale_y_log10(limits = c(1e-2, NA)) +
  scale_x_discrete(limits = x_axis_order) +
  scale_color_manual(values = binary_colors) +
  plot_theme +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "Distrubution of Chunk Coverage by Bin (Boxplot)",
    subtitle = paste0("Chunk size: ", chunk_size),
    x = "Bin (Sorted by median depth of nondigest)",
    y = "Chunk Coverage (log10)"
  )
```

## Cleaner distrubution of chunks
The same plot is generated again Figure \@ref(chunkQuantilePlot), but only with Q25 to Q75 to more easily see the differences between the samples.
Additionally, Wilcox rank sum test is used to check p-value of the difference between before and after digestion (bonferroni corrected).

```{r chunkQuantilePlot}
merge(chunks_stats, chunks_sampleVSbin, by = "bin_char") %>%
  ggplot() +
    aes(x = bin_char, y = median, color = sample) +
    geom_hline(yintercept = 1, linetype = 1, color = "black") +
    geom_errorbar(aes(ymin = q_min, ymax = q_max), width = 0.3) +
    geom_point(aes(fill = p_val), shape = 21, size = 1.5, color = "#000000") +
    scale_y_log10(limits = c(1e-2, NA)) +
    scale_x_discrete(limits = x_axis_order) +
    scale_color_manual(
      values = binary_colors,
      guide = guide_legend(
        title = "Sample",
        title.position = "top",
        title.hjust = 0.5
      )
    ) +
    scale_fill_continuous(
      low = "#000000", high = "white",
      na.value = "#000000",
      trans = "log10", limits = c(1e-6, 1),
      guide = guide_colorbar(
        title = "p-value (Wilcox)",
        label.theme = element_text(angle = 30, size = 8),
        title.position = "top",
        label.hjust = 1,
        reverse = TRUE,
        title.hjust = 0.5
      )
      ) +
    plot_theme +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      title = "Chunk Coverage by Bin (Q25-Median-Q75)",
      subtitle = paste0("Chunk size: ", chunk_size),
      x = "Bin",
      y = "Chunk Coverage (log10)",
      fill = "p-value",
      color = "Sample"
    )
```


## Initial remarks
Bins seem to be targeted differentially, some are greatly reduced in chunk cvoerages and other are barely depleted.
In figure \@ref(chunkQuantilePlot) the IQR is also quite nice for many of chunk digest chunk coverages, meaning atleast 50% of the bin is stably depleted. 


# Within bins spread
## Difference in IQR
The change in chunk coverage inter quarttile range before and after digestion is plotted.
The direction of the arrow is the direction of the change from nondigest to digest, and the length of the arrow is the change in inter quarttile range.
Generally, it seems the digest chunk have a smaller inter quarttile range. This is counterintutive, and is probabl due to increase in IQR as mean increase. 

```{r}
x_axis_IQR_plot <- chunks_stats[
  sample == "Nondigest",
  ][
    order(IQR, decreasing = TRUE),
    bin_char
  ]
dcast(chunks_stats, bin_char ~ sample, value.var = c("IQR", "mean")) %>%
  ggplot() +
    aes(x = bin_char) +
    geom_segment(
      aes(
        y = IQR_Nondigest,
        yend = IQR_MboI_digest,
        xend = bin_char,
        color = fifelse(IQR_Nondigest > IQR_MboI_digest, "Nondigest", "MboI_digest")
      ),
      size = 1,
      arrow = arrow(length = unit(0.1, "cm"))
    ) +
    scale_x_discrete(limits = x_axis_IQR_plot) +
    plot_theme +
    scale_y_continuous(expand = c(0.005, 0)) +
    scale_color_manual(values = binary_colors) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      title = "IQR of Chunk Coverages by Bin",
      subtitle = paste0("Chunk size: ", chunk_size),
      x = "Bin (Sorted by nondigested)",
      y = "Inter Quartile Range (IQR)",
      color = "Highest IQR"
    )
```

## IQR correlation to Mean
As expected IQR increases as mean increases (Figure\@ref(SDvsMeanBinCoverage)) .
The change of each bin is also indicated by the connecting line. 
The change of depleted bins is generally towards lower mean and a little lower IQR. 


```{r SDvsMeanBinCoverage}
chunks_stats %>%
ggplot() +
  aes(y = IQR, x = mean, color = sample) +
  geom_segment(
    data = dcast(chunks_stats, bin_char ~ sample, value.var = c("IQR", "mean")),
    mapping = aes(
      x = mean_Nondigest,
      y = IQR_Nondigest,
      xend = mean_MboI_digest,
      yend = IQR_MboI_digest
    ),
    inherit.aes = FALSE,
    alpha = 0.2
  ) +
  geom_point() +
  scale_y_log10() +
  scale_x_log10() +
  scale_color_manual(values = binary_colors) +
  geom_smooth(method = "lm", se = FALSE, size = 2, ) +
  labs(
    title = "Bin Coverage inter quarttile range vs Mean",
    subtitle = paste0("Chunk size: ", chunk_size),
    x = "Mean Bin Coverage (log10)",
    y = "inter quarttile range of Bin Coverage (log10)",
    color = "Sample"
  ) +
  plot_theme +
  theme(
    legend.position = "bottom"
  )
```

## Change in IQR and Mean
The change in IQR and mean (essentially the connecting line Figure \@ref(SDvsMeanBinCoverage)) is 
plotted in Figure \@ref(SDvsMeanBinCoverageChange).
The change in correlates very nicely with the change in mean.

```{r}
dcast(chunks_stats, bin_char ~ sample, value.var = c("IQR", "mean"))[
    , IQR_diff := IQR_Nondigest - IQR_MboI_digest
  ][
    , mean_diff := mean_Nondigest - mean_MboI_digest
  ] %>%
  ggplot() +
    aes(x = mean_diff, y = IQR_diff) +
    geom_point() +
    scale_y_log10() +
    scale_x_log10() +
    labs(
      title = "Bin Coverage inter quarttile range diff vs Mean diff",
      subtitle = paste0("Chunk size: ", chunk_size),
      x = "IQR diff\nNondigest - MboI Digest",
      y = "Mean diff\nNondigest - MboI Digest",
      color = "Sample"
    ) +
    plot_theme

```

To investigate the difference in spread between the samples, the coverage is scaled by the mean coverage of the bin.
This is done eliminate the increase in values that arise at higher coverage values. 
The scaling happens within each bin and sample, so the mean of the scaled values is 1 within groups.



```{r chunkBoxplotScaledMean}
ggplot(chunks) +
  aes(x = bin_char, y = depth_scaled_mean, color = sample) +
  geom_boxplot(outlier.alpha = 0, outlier.size = 0.1) +
  scale_x_discrete(limits = x_axis_order) +
  scale_color_manual(values = binary_colors) +
  ylim(-1, 4) +
  plot_theme +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "Difference in chunk spread (coverage / mean(coverage))",
    subtitle = paste0("Chunk size: ", chunk_size),
    x = "Bin (Sorted by median depth of nondigest)",
    y = "Scaled Chunk Coverage"
  )
```

At a glance the spread of the digested samples appear larger. 
The shift in inter quarttile range and IQR of scaled means is investigated to evaluate this \@ref()

```{r SDplotScaledMean}
x_axis_IQR_plot <- chunks_stats_scaled_mean[
  sample == "Nondigest",
  ][
    order(IQR, decreasing = TRUE),
    bin_char
  ]
dcast(chunks_stats_scaled_mean, bin_char ~ sample, value.var = "IQR") %>%
  ggplot() +
    aes(x = bin_char) +
  scale_color_manual(values = binary_colors) +
    geom_segment(
      aes(
        y = Nondigest,
        yend = MboI_digest,
        xend = bin_char,
        color = fifelse(Nondigest > MboI_digest, "Nondigest", "MboI_digest")
      ),
      size = 1,
      arrow = arrow(length = unit(0.1, "cm"))
    ) +
    scale_x_discrete(limits = x_axis_sd_plot) +
    plot_theme +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      title = "IQR of Scaled Chunk Coverages by Bin",
      subtitle = paste0("Chunk size: ", chunk_size),
      x = "Bin (Sorted by nondigested IQR)",
      y = "Inter Quartile Range (IQR)",
      color = "Highest IQR"
    )
```
